---
title: 'Gap Minder - Problem Set 1 Lesson 4, #15'
output:
  pdf_document: default
  html_notebook: default
---

Execute with *Run* button within the chunk or by placing your cursor inside it and pressing *Cmd+Shift+Enter*. 
Add  new chunk with *Insert Chunk* button on toolbar or *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file).

###The Gapminder website
contains over 500 data sets with information about the world's population. Your task is to download a data set of your choice and *create 2-5 plots* that make use of the techniques from Lesson 3.
_You might use:_
- a simple histogram, 
- a boxplot split over a categorical variable,
- or a frequency polygon
The choice is yours!

###_Once you've completed your investigation, create a post in the discussions that includes:_
1. any questions you answered, your observations, and summary statistics
2. snippets of code that created the plots
3. links to the images of your plots

###Save images by using the ggsave() command.
ggsave() will save the last plot created.
 For example...
#                  qplot(x = price, data = diamonds)
#                  ggsave('priceHistogram.png')

ggsave currently recognises the extensions eps/ps, tex (pictex),
pdf, jpeg, tiff, png, bmp, svg and wmf (windows only).

###Read Files: _`read.csv()`_
The following command may be helpful for some of the Gapminder Data sets, once it's been converted to csv format: `read.csv('data.csv', header = T, row.names = 1, check.names = F)`. You may want to look at additional function arguments in the help file for the `read.table()` family for additional tools that can help you read in data.

###To exchange the rows and columns of  dataframe 
 use the transpose function, `t()`.
 
###*Copy and paste all of the code that you used for your investigation, and submit it when you are ready.*
---
####Read in File
```{r ReadData}
unemployed <- read.csv('f 25-54 unemploy.csv', header=TRUE)
#ingore last row 30 and column (27) with NA values
unplyd <- unemployed[1:29,1:26]
names(unplyd)<- gsub('X', "", names(unplyd))
colnames(unplyd)[1] = "country"
```
####Load Libraries
```{r LoadLibraries}
library(tidyr)
library(dplyr)
library(ggplot2)
library(gridExtra)
```
####Tidy Data

collapse multiple columns with gather
```{r TidyData}

unplyd_year <- gather(unplyd, year, unemployed, 2:26,
                      na.rm = TRUE)
head(unplyd_year,5)
str(unplyd_year)
```
###Data Transformation with dplyr
filter for unemployment > 20%
arrange from largest to smallest (descending) unemployment
arrange from most recent to oldest year (descending)
```{r}

big_u <- filter(unplyd_year, unemployed >20)
big_u
arrange(big_u, desc(unemployed))
```
```{r Boxplot_plain}
boxplot(unplyd_year$unemployed~unplyd_year$country, 
        ylab = '% Unemployment', las=2)
```
###Group/Order countries by median unemployment
```{r}
bymedian <- with(unplyd_year,reorder(unplyd_year$country,unplyd_year$unemployed,median))
bymedian
```

```{r MedianUnemployment}
boxplot(unplyd_year$unemployed~bymedian, 
        ylab= "% Unemployment", las=3,
        par(mar = c(7, 5, 1, 2) + 0.1), col = 'lightblue',              cex.axis = 0.7,
        cex.lab = 0.9)
mtext('World Survey Unemployed Females aged 25-54', side=3, line = 0)
mtext('Country', side=1, line = 6)
text(9,10,"US",cex= 0.6, col= 'red')

```
###Plot world unemployment by year
```{r}
boxplot(unplyd_year$unemployed~unplyd_year$year,
        las=2, cex.axis = 0.8, col= 'lightgreen')
mtext('World Survey Unemployed Females aged 25-54', side=3, line = 1)
mtext('% Unemployment', side=2, line = 2)
mtext('Year', side=1, line = 3)
```
###Count = number of countries reporting results
```{r}
qplot(x = unplyd_year$year, data = unplyd_year, 
      fill = I('darkred'),
      xlab = 'Year', ylab = 'Count',
      main = '# Countries Reporting Unemployment',
      )+
   theme(plot.title = element_text(hjust = 0.5), axis.text.x = element_text(angle = 90, hjust = 0))

```
```{r}
unplyd_year %>% group_by(year) %>% summarise(n=n())
```


###Statistics of Unemployment by Year
```{r}
by(unplyd_year$unemployed,unplyd_year$year, summary)
```
###Order by Countries, then Year
```{r}
#using pipe %>%
acountry <- unplyd_year %>% arrange(country, year)
acountry
#using group_by
grouped <-group_by(acountry, country, year)
```
###Line graph of Unemployment by year in all Countries
```{r}
ggplot(data=acountry, aes(acountry$year, acountry$unemployed,
                          colour = acountry$country,
                          group = acountry$country))+
  geom_line()+
  xlab('YEARS') + 
  ylab('% Unemployed')+
  theme(axis.text.x=element_text(angle=60, hjust=1), 
        legend.key.size = unit(0.1,'cm'),
        legend.text = element_text(angle = 10, size = 6),
        legend.title = element_blank())+
  ggtitle('US Unemployed Females aged 25-54')
```
###Summarise Stats by group (country or year)
cannot skip first pipe otherwise grouping is wrong
```{r}
acountry %>% 
  group_by(country) %>%
  summarise(mean=mean(unemployed), sum = sum(unemployed),
  n= n())

acountry %>% 
  group_by(year) %>%
  summarise(mean=mean(unemployed), sum = sum(unemployed),
  n= n())

# following gives overall stats for entire DF not "arranged" dataframe
acountry %>% 
  summarise(mean=mean(unemployed), sum = sum(unemployed),
  n= n())
```

###Graph of 2 countries from separate DFs
```{r}
us <-subset(acountry, acountry$country == 'United States')
spain <-subset(acountry, acountry$country == 'Spain')
ggplot()+
  geom_line(aes(us$year, us$unemployed, 
                color =country, group = 1),us)+
  geom_line(aes(spain$year, spain$unemployed, 
                color =country, group =1),spain) + 
  xlab('YEARS') + 
  ylab('% Unemployed')+
  theme(axis.text.x=element_text(angle=60, hjust=1))+
  ggtitle('Unemployed Females aged 25-54')
```
###Distribution of the Unemployment Rate
```{r}
dist1 <-qplot(x = unplyd_year$unemployed, data = unplyd_year, binwidth = 1,
      xlab = "Unemployment Rates of the World 1981-2005",
      ylab = "Count",
      color = I('black'), fill = I('chocolate3'))+
  scale_x_continuous(breaks = seq(0,25, by=5), limits = c(0,25))
dist1
```
###Frequency Polygon 
add frequency line to distribution graph
```{r}
dist1 + geom_freqpoly(colour = 'blue', binwidth = 0.75)
```
###Transformations
Square Root transformation of scale
```{r}

dist2 <-qplot(x = unplyd_year$unemployed, data = unplyd_year, binwidth = .2,
      xlab = "Unemployment Rates of the World 1981-2005",
      ylab = "Count",
      color = I('black'), fill = I('blue'))+
  scale_x_sqrt(breaks = seq(0,30, by=5), limits = c(0,35))
dist2
```
###Facet Wrap
first option does some type of couont/distribution plot while the 
second with geom_line plots the unemployment v year per country
```{r}
qplot(unplyd_year$unemployed, 
      data =unplyd_year,
      binwidth = 1) +
  scale_x_continuous(breaks = 1:31) +
  facet_wrap(~unplyd_year$country, scales ='free')
```
```{r}
ggplot(unplyd_year, aes(year, unemployed, group=country))+
  geom_line()+
  facet_wrap(~unplyd_year$country, scales ='free')
```

```{r}
subset(unplyd_year, unplyd_year$country == 'Spain')
```
```{r}
sample(unplyd_year$country,5)
```

```{r}
unplyd_year[sample(nrow(unplyd_year), 5),]
```

```{r}
ggplot(acountry, aes(acountry$unemployed, 
                     acountry$country,))+
  geom_point()
```

```{r}
acountry2 <- acountry
unique(levels(acountry2$country))
levels(acountry$country)
```
###Relevel the countries by unemployment
```{r}
acountry2$country2 <-factor(
  acountry$country,
  levels=acountry2[order(acountry2$unemployed), 'country'])
levels(acountry2$country)
```
```{r}
unique(levels(acountry2$country))
```

```{r}

ggplot(acountry2, aes(y=country2, x=unemployed))+
  geom_point(stat="identity")

```
```{r}
mittel <-acountry2 %>% group_by(country) %>% 
  summarise(mean=mean(unemployed))

a<-ggplot(mittel, aes(y=mittel$country, x=mittel$mean))+
  geom_point()

ag_country <- aggregate(unemployed ~ country, acountry, max)

b<- ggplot(ag_country, aes(y=ag_country$country,
                       x=ag_country$unemployed))+
  geom_point()

grid.arrange(a,b, nrow=2)
```

```{r}
acountry2$country3 <-factor(
  acountry2$country, levels=ag_country[
    order(ag_country$unemployed), "country"])

ggplot(acountry2, aes(y=country3, x=unemployed))+
  xlab('Unemployment %')+
  ylab('Country')+
  geom_point(stat="identity")+
  ggtitle("Unemployment Organized by mean/max/min")
```
###There is a new Ending 
